Variance reduction for MCMC methods via martingale representations

### 2020-02-18 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex
update

\def\PE{\mathsf{E}}
\newcommandx{\CPE}[3][1=]{{\mathsf E}_{#1}\left[\left. #2 \, \right| #3 \right]}
\usepackage[textsize=footnotesize]{todonotes} % pour voir les commentaires

@@

%added command for indicator - SS
\newcommand{\indiacc}[1]{\mathbbm{1}_{\{#1\}}}

@@

\newcommand{\tcr}[1]{\textcolor{red}{#1}}

\begin{document}

@@

\eric{do we need a trace here ?}

@@

\eric{I would write \tcr{$L^2(P_\xi)$} to be coherent with the previous notations} 

@@

Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq l,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.

@@

should be small for \(K\) large enough. Indeed, due to orthonormality  we have

@@

We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.

@@

\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
+\mathsf{Var}[\rho(f)].
Our analysis will be done under the following two assumptions.

@@

\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L_U\geq 0$ such that

@@

|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \mathbb{R}^d.

@@

\langle D^2U(x),x\rangle \geq \frac{m_U}{2}\|x\|^2.

@@

\subsection{Squared bias}
Due to the martingale transform structure

@@

with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$ For a bounded Borel function~$f$,

@@

similarly to the beginning of ~\cite[Section~4]{durmus:moulines:2017}:
\left(\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
&=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
\label{eq:06042018a2}\\
&\le

@@

\|\delta_x Q_\gamma^{p}-\pi(\cdot)\|_{TV}^2,

@@

$\|\mu-\nu\|_{\TV}$ denotes the total variation distance

@@

\|\mu-\nu\|_{TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
|\mu(A)-\nu(A)|,
$\pi(\cdot)$ denotes the probability measure on $\mathbb R^d$
for $\gamma>0$, the Markov kernel $Q_\gamma$
is defined by
\mathcal{Q}_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy.

@@

on bounds in the total variation distance. It is known  that under assumptions {\bf (H1)} and {\bf (H2)}  the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. Moreover, as shown in Proposition (Appendix) \begin{eqnarray}

@@

\|\delta_x\mathcal{Q}_{\gamma}^{n}-\pi_\gamma\|_{TV} \leq C_1\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right), \quad \|\pi-\pi_\gamma\|_{TV} \leq C_2 \sqrt{\gamma}

@@

\subsection{Variance}
An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}

@@

\mathsf{Var}[\pi_n^N(f)]\lesssim \frac{1}{n\gamma},

@@

@@

|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} \mu (x)|\leq B_\mu, \quad

@@

@@

\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{\gamma^{K-2}}{n},

@@

where \(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)

@@

where $F_s(X_0,Z_1,\ldots,Z_s)=\frac{1}{s}\sum_{r=0}^s f(X_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$

@@

\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)

@@

\mathbf{\mathbf{K}}_{I}=(K1_{\{1\in I\}},\ldots,K1_{\{d\in I\}}).

@@

\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)

>>>>

\def\PE{\mathsf{E}}
\def\PVar{\mathsf{Var}}
\newcommandx{\CPE}[3][1=]{{\mathsf E}_{#1}\left[\left. #2 \, \right| #3 \right]}
\newcommand{\ps}[2]{\langle #1, #2 \rangle}
\def\nset{\mathbb{N}}
\def\rset{\mathbb{R}}
\usepackage[textsize=footnotesize]{todonotes} % pour voir les commentaires

@@

%added command for indicator - SS
\newcommand{\indi}[1]{\mathbbm{1}_{#1}}
\newcommand{\indiacc}[1]{\mathbbm{1}_{\{#1\}}}

@@

\newcommand{\tcr}[1]{\textcolor{red}{#1}}
\newcommand{\tvnorm}[1]{\| #1 \|_{\operatorname{TV}}}
\def\kerULA{\operatorname{Q}}
\def\eqsp{\,}
\begin{document}

@@

\eric{do we need a trace here ? I think also that $D^2$ is not defined }

@@

\eric{I would write \tcr{$L^2(\mathsf{P})$} to be coherent with the notation $\mathsf{E}$ and we need to define what is the underlying probability space}

@@

Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 1,\)} \eric{I do not understand this}  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.

@@

should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_k'(\xi_l)]=0$ if $k \ne k'$, we obtain

@@

\tcr{We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.} \eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}

@@

\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
Our analysis is carried out under the following two assumptions:

@@

\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L_U < \infty$ such that

@@

|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \rset^d.

@@

\ps{D^2U(x)}{x} \geq (m_U/2)\|x\|^2.

@@

\paragraph{Squared bias:} Due to the martingale transform structure

@@

with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$

Denote by $\pi($ denotes the probability measure on $\mathbb R^d$
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one-step of the ULA algorithm by
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.
\eric{what is the notation convention here; I do not like much $\mathcal{Q}$ but $Q$ is already define and it is perhaps not a great idea to supersede the previous notation}
For a bounded Borel function~$f$,

@@

similarly to~\cite[Section~4]{durmus:moulines:2017}:
\label{eq:06042018a2}
\left\{\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right\}^2
=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
\le

@@

\tvnorm{\delta_x \kerULA_\gamma^{p}-\pi(\cdot)}^2,

@@

$\tvnorm{\mu-\nu}$ denotes the total variation distance

@@

\tvnorm{\mu-\nu}=\sup_{A\in\mathcal B(\mathbb R^d)}
|\mu(A)-\nu(A)| \eqsp.

@@

on bounds in the total variation distance. It is known  that under assumptions {\bf (H1)} and {\bf (H2)},   the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. Moreover, as shown in Proposition, \tcr{(Appendix)}
\eric{put a precise reference here} 

@@

\tvnorm{\delta_x\kerULA_{\gamma}^{n}-\pi_\gamma} \leq C_1\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right), \quad \tvnorm{\pi-\pi_\gamma} \leq C_2 \sqrt{\gamma}

@@

\paragraph{Variance:} An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}

@@

\PVar[\pi_n^N(f)]\lesssim (n\gamma)^{-1} \eqsp,

@@

\erici{the notation $\partial^{\mathbf{k}}$ should be defined here}

@@

|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} U (x)|\leq B_\mu, \quad

@@

\eric{I have replaced $\mu$ by $U$; in this section we use mostly $U$ and not $\mu$; we could use $\mu$ but in such case it would be better to recall that $\mu= \nabla U/2$}
\eric{Is it required to assume that the gradient is bounded ? It contradicts assumption (H2) which assumes a quadratic growth at infinity - coercivity of the operator ?}

@@

\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-2} \eqsp,

@@

where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
\eric{this is defined less precisely just above}

@@

where $F_s(X_0,Z_1,\ldots,Z_s)=s^{-1}\sum_{r=0}^s f(X_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$

@@

\PVar_x\left( s^{-1} \sum\nolimits_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)

@@

\mathbf{\mathbf{K}}_{I}=K(\indi{I}(1),\ldots, \indi{I}(d)).

@@

 \eric{Since we do note state was $W_s$ is, it is a bit difficult to guess why we are using a Poincare inequality there}
\PVar_x\left(s^{-1} \sum\nolimits_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
